Type theories as quotient inductive-recursive types

### 2025-02-06 tex/main.tex
polish Section 3

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume uniqueness of identity proofs globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{X}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil & : \JList{X} \\
  \Jsing{-} & : X \to \JList{X} \\
  \Jconcat{-}{-} & : \JList{X} \to \JList{X} \to \JList{X} \\
  \mathsf{idl} & : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{X}} \ell \\
  \mathsf{idr} & : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{X}} \ell \\
  \mathsf{assoc} & : \implicit{\ell_0, \ell_1, \ell_2}\; (\Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2}) =^{\JList{X}} (\Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})})
\end{alignat*}

@@

\Fredrik{Give an actual example of QIIT (or even QIRT)??}

Shulman~\cite{Shulman2014} considered combining higher inductive types and inductive-recursive types, but the foundational status of this combination is underdeveloped. We will also make use of such a combination, which gives additional motivation to study its metatheory in future work.

@@

We begin with the definition of (parallel) substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
Then we introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
Then, we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types under some context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from the domain $\Gamma$ and the codomain $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms under a context $\Gamma$ and its type $A$. 
These types amount to the following types indexed by types being defined (hence inductive-inductive):
\begin{alignat*}{3}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
The type $\Ctx$ has two constructors $\emptyctx$ for the empty context and $\blank,\blank$ for context extension:
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
where the context extension $\blank,\blank$ requires that $\Ty$ is indexed by a context $\Gamma$, underscoring the nature of inductive-inductive definition.

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.

If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
Hence another constructor is introduced:
\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
\end{alignat*}
For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families), 
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
satisfying certain laws (introduced later as equality constructors).
Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$
for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution:
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
\end{alignat*}
such that $[\sigma]\,t$ is of $\Tm\;\Gamma\;([\sigma]\,A)$.
Note that $[\blank]\blank$ is overloaded for type and term substitution.

@@

The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the length of terms in $\sigma$, giving rise two $\eta$-laws in the case that $\Delta$ is empty and non-empty:
\begin{alignat*}{5}
  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma} && \sigma        && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma} && \sigma   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma)
\end{alignat*}

The laws for substitution composition are stipulated using equality constructors.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
\end{alignat*}
Substitution composition $\blank;\blank$ acts like a list concatenation, so $\sigma; (\tau, t)$ is intuitively equal to $(\sigma; \tau), [\sigma]\,t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
However, by the type of substitution extension, $[\sigma]\,t$ requires to be of type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead, so we have to transport $[\sigma]\,t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.

The tail and the head of a non-empty substitution $(\sigma, t)$ is obviously $\sigma$ and $t$ respectively:
\begin{alignat*}{5}
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
\end{alignat*}
whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
Therefore, we have to transport $\pi_2(\sigma, t)$ along $\pi_1\beta$, so the above equality constructor is, in fact, $\transfib{\Tm\;\Gamma\;([\blank]\,A)}{\pi_1\beta}{\pi_2(\sigma, t)} =^{\Tm\;\Gamma\;([\sigma]\,A)} t$.

Similarly, for term substitution, terms are transported along the corresponding rules:
\begin{alignat*}{5}
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
\end{alignat*}
equivalent to the following (homogeneous) identities
\[
  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
  \quad\text{and}\quad
  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
\]
respectively.
Constructors introduced so far complete the definition of substitution calculus.

@@

To retain the intuitive way of reasoning such as \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution as a constructor does not compute so that type mismatches occur with substitution.

One possibility is to define substitution calculus as a QII`R' type where type substitution, as a simultaneously defined recursion, does compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
This definition fails to remove any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching.
\LT[noinline]{How about \Coq and \Lean?}
In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

>>>>

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume uniqueness of identity proofs globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{X}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil & : \JList{X} \\
  \Jsing{\blank} & : X \to \JList{X} \\
  \Jconcat{\blank}{\blank} & : \JList{X} \to \JList{X} \to \JList{X} \\
  \mathsf{idl} & : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{X}} \ell \\
  \mathsf{idr} & : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{X}} \ell \\
  \mathsf{assoc} & : \implicit{\ell_0, \ell_1, \ell_2}\; (\Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2}) =^{\JList{X}} (\Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})})
\end{alignat*}

@@

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{X}$ is defined simultaneously with a recursively defined function $\flatten : \JList{X} \to \List{X}$ which flattens a join list into an ordinary list. We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
\mathsf{nf} : (\ell, \ell' : \JList{X}) \to \flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell' \to \ell =^{\JList{X}} \ell'
\]
which states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{X}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$.

%Shulman~\cite{Shulman2014} considered combining higher inductive types and inductive-recursive types, but the foundational status of this combination is underdeveloped. We will also make use of such a combination, which gives additional motivation to study its metatheory in future work.

@@

We begin with the definition of the parallel substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
Then, we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

Substitution calculus has a type $\Ctx$ of contexts, a type $\Ty\,\Gamma$ of types in context $\Gamma : \Ctx$, a type $\Sub\;\Gamma\;\Delta$ of substitutions from context $\Gamma$ to context $\Delta$, and a type $\Tm\;\Gamma\;A$ of terms in context $\Gamma$ of type $A$. 
This amounts to the following type families, all mutually inductively defined (hence inductive-inductive):
\begin{alignat*}{3}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
The type $\Ctx$ has two constructors: $\emptyctx$ for the empty context, and $\blank,\blank$ for context extension, typed as follows.
\begin{alignat*}{3}
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
Note that a context extension $\Gamma , A$ is only wellformed if $A : \Ty\;\Gamma$, ensuring overall wellformedness of contexts, but requiring contexts and types to be defined simultaneously.

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ are kept implicit.
\Fredrik{Could we change this to $A[\sigma]$ instead?}

Since $\Ty$ is inductively defined, we add a base case to ensure that $\Ty\;\Gamma$ is not empty for every $\Gamma$:
%
%If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
%For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
%Hence another constructor is introduced:
\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
\end{alignat*}
For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$, because the substitution calculus is supposed to be the initial \emph{category} with families:
%Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families),
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
These need to satisfy certain laws, which we introduce later as equality constructors.
Context extension is characterised by its projections, which picks out the rest of the context and the most recent variable
%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution, which given a substitution $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in context $\Delta$ to terms in context $\Gamma$, appropriately preserving the type, which is also substituted:
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
\end{alignat*}
Note that $[\blank]\blank$ is overloaded for type and term substitution.

@@

The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the number of terms in $\sigma$, giving rise to two $\eta$-laws --- one for $\Delta$ being empty, and one for $\Delta$ being non-empty:
\begin{alignat*}{5}
  \emptyctx\eta   & : \implicit{\Gamma, A, \sigma}\; && \sigma        && =^{\Sub\,\Gamma\,\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, \Delta, \sigma}\; && \sigma   && =^{\Sub\,\Gamma\,(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma)
\end{alignat*}

The laws for substitution composition are stipulated using equality constructors, where, for the first time, we see the need to adjust the types involved with a $\transp$ in order for both sides to have the same type.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
\end{alignat*}
%
Composition of substitutions $\blank;\blank$ acts like a list concatenation, so
we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), [\sigma]\,t$. However $[\sigma]\,t$ is of type $\Tm\;\Gamma\;([\sigma]\,[\tau]\;A)$, but the context extension $(\sigma; \tau), [\sigma]\,t$ requires $[\sigma]\,t$ to have type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead. Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ which we can transport along to fix up the equation.

%$\sigma; (\tau, t)$ is intuitively equal to $(\sigma; \tau), [\sigma]\,t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
%However, by the type of substitution extension, $[\sigma]\,t$ requires to be of type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead, so we have to transport $[\sigma]\,t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.

The first and second projections of a non-empty substitution $(\sigma, t)$ should  obviously be $\sigma$ and $t$ respectively
\begin{alignat*}{5}
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
\end{alignat*}
but again, note that the type of $\pi_2\beta$ only makes sense because we can transport along $\pi_1\beta$ to make sure both sides of the equation have the same type: $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$.
%whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
%Therefore, we have to transport $\pi_2(\sigma, t)$ along $\pi_1\beta$, so the above equality constructor is, in fact, $\transfib{\Tm\;\Gamma\;([\blank]\,A)}{\pi_1\beta}{\pi_2(\sigma, t)} =^{\Tm\;\Gamma\;([\sigma]\,A)} t$.
%
Similarly, for term substitution, terms are transported along the corresponding rules:
\begin{alignat*}{5}
  [\idS]_t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}         && =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}}  && t \\
  [;]_t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}   && {[ \sigma ]\,[ \tau ]\,t}
\end{alignat*}
equivalent to the following (homogeneous) identities
\[
  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
  \quad\text{and}\quad
  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
\]
respectively.
Constructors introduced so far complete the definition of substitution calculus.

@@

To retain the intuitive way of reasoning such as in equation \eqref{eq:pi2-comp-proof} \emph{formally}, we would like to make type substitution rules definitional, since the root cause is that type substitution does not compute since it is a constructor, leading to type mismatches occurring with substitution.

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
\LT[noinline]{How about \Coq and \Lean?}
\Fredrik[noinline]{They also translate to eliminators, so I think we do not need to say anything}
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the definitional equality $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.
